---
title: "Summarise population characteristics"
output: 
  html_document:
    pandoc_args: [
      "--number-offset=1,0"
      ]
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{C-summarise_pop_characteristics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In this vignette, we will explore the *OmopSketch* functions that provide information about individuals characteristics at specific points in time. We will employ `summarisePopulationCharacteristics()` to generate a summary of the demographic details within the database population. Additionally, we will tidy and present the results using `tablePopulationCharacteristics()`, which supports either [gt](https://gt.rstudio.com/) or [flextable](https://davidgohel.github.io/flextable/) for formatting the output.

## Create a mock cdm

Before we dive into *OmopSketch* functions, we need first to load the essential packages and create a mock CDM using the function `mockOmopSketch()`.

```{r, warning=FALSE}
library(dplyr)
library(OmopSketch)


cdm <- mockOmopSketch()
```

# Summarise population characteristics
To start, we will use `summarisePopulationCharacteristics()` function to generate a summarised result object, capturing demographic characteristics at both `observation_period_start_date` and `observation_period_end_date`.


```{r, warning=FALSE}
summarisedResult <- summarisePopulationCharacteristics(cdm)

summarisedResult |> glimpse()
```
To tidy and display the summarised result using a [gt](https://gt.rstudio.com/) table, we can use `tablePopulationCharacteristics()` function.
```{r, warning=FALSE}
summarisedResult |>
  tablePopulationCharacteristics(type = "flextable")
```
To obtain a [flextable](https://davidgohel.github.io/flextable/) instead of a [gt](https://gt.rstudio.com/), you can simply change the `type` argument to `"flextable"`. Additionally, it is important to note that age at start, prior observation, and future observation are calculated at the start date defined (in this case, at individuals observation_period_start_date). On the other hand, age at end is calculated at the end date defined (i.e., individuals observation_period_end_date). 

## Trim study period
To focus on a specific period within the observation data, rather than analysing the entire individuals' observation period, we can trim the study period by using the `studyPeriod` argument. This allows to analyse the demographic metrics within a defined time range rather than the default observation start and end dates.
```{r, warning=FALSE}
summarisePopulationCharacteristics(cdm,
                                   studyPeriod = c("1950-01-01", "1999-12-31")) |>
  tablePopulationCharacteristics()
```

However, if you are interested in analysing the demographic characteristics starting from a specific date without restricting the study end, you can define just the start of the study period. By default, `summarisePopulationCharacteristics()` function will use the observation_period_end_date to calculate the end-point statistics when the end date is not defined. 

```{r, warning=FALSE}
summarisePopulationCharacteristics(cdm,
                                   studyPeriod = c("1950-01-01", NA)) |>
  tablePopulationCharacteristics()
```

Similarly, if you are only interested in analysing the population characteristics up to a specific end date, you can define only the end date and set the `startDate = NA`. By default the observation_period_start_date will be used.

## Stratify by age groups and sex
Population characteristics can also be estimated by stratifying the data based on age and sex using `ageGroups` and `sex` arguments.
```{r, warning=FALSE}
summarisePopulationCharacteristics(cdm,
                                   sex = TRUE,
                                   ageGroup = list("<60" = c(0,59), ">=60" = c(60, Inf))) |>
  tablePopulationCharacteristics()
```

