---
title: "Summarise the person table"
output: 
  html_document:
    pandoc_args: [
      "--number-offset=1,0"
      ]
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{summarise_person}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette walks through the **OmopSketch** functions designed to provide a concise overview of the OMOP `person` table. Two functions cover this workflow:

- `summarisePerson()`: computes summary statistics and data-quality checks for the `person` table, including total subject counts, observation-period coverage, sex/race/ethnicity distributions, birth-date components, and summaries of id-columns (`location_id`, `provider_id`, `care_site_id`).
- `tablePerson()`: renders the results as a formatted table (`gt`, `reactable`, or `datatable`).

## Setup

Load the required packages and create a mock CDM using [omock](https://ohdsi.github.io/omock/).

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(OmopSketch)
library(omock)

cdm <- mockCdmFromDataset(datasetName = "GiBleed", source = "duckdb")
cdm
```

# Summarise the person table

Call `summarisePerson()` to compute all summaries. It returns a
[`summarised_result`](https://darwin-eu.github.io/omopgenerics/articles/summarised_result.html)
object — a standardised tidy format used across the OMOP analytics ecosystem.

```{r}
result <- summarisePerson(cdm = cdm)
result |> glimpse()
```

## What the function reports

`summarisePerson()` covers the following summaries, each stored as a separate
`variable_name` in the result:

| Variable | Description |
|---|---|
| **Number subjects** | Total row count in `person`. |
| **Number subjects not in observation** | Count and percentage of persons absent from `observation_period`. A warning is emitted when this is non-zero. |
| **Sex** | Count and percentage for `Female`, `Male`, and `None` (derived via `PatientProfiles::addSexQuery()`). |
| **Sex source** | Distribution of raw `gender_source_value`. |
| **Race** | Distribution of `race_concept_id`, resolved to concept names. |
| **Race source** | Distribution of raw `race_source_value`. |
| **Ethnicity** | Distribution of `ethnicity_concept_id`, resolved to concept names. |
| **Ethnicity source** | Distribution of raw `ethnicity_source_value`. |
| **Year of birth** | Numeric summary: missingness, quantiles (Q05, Q25, median, Q75, Q95), min/max. |
| **Month of birth** | Same numeric summary as year of birth. |
| **Day of birth** | Same numeric summary as year of birth. |
| **Location** | Missing count, zero count, and distinct values for `location_id`. When `location_id` is empty, the function attempts to derive it from `care_site_id` and notes this in a message.|
| **Provider** | Missing count, zero count, distinct values, and (below threshold) individual `provider_name` labels. |
| **Care site** | Missing count, zero count, distinct values, and (below threshold) individual `care_site_name` labels. |

### Threshold

For `location_id`, `provider_id`, and `care_site_id`, `summarisePerson()` always
reports three aggregate statistics: number missing, number of zeros, and number
of distinct values. Whether it goes further and lists each value individually
depends on a threshold.

If the number of distinct values is **below 15** (the default), the function
joins to the corresponding lookup table and appends one row per unique label —
`location_source_value`, `provider_name`, or `care_site_name` — each with its
own count and percentage. This gives you a readable breakdown rather than just a
count of how many distinct ids exist.

If the number of distinct values is **at or above 15**, only the aggregate stats
are returned. Listing dozens or hundreds of individual sites would produce noise
rather than insight.

You can tune this cutoff for your database before calling `summarisePerson()`:

```{r, eval=FALSE}
# Raise the threshold to show individual labels for up to 50 distinct values
options(OmopSketch.personLabels = 50)
result <- summarisePerson(cdm = cdm)
```

### Estimate types

Depending on the variable, estimates include:

- `count` / `percentage` — for categorical variables (sex, race, ethnicity).
- `count_missing` / `percentage_missing` — missingness for id-columns and birth-date fields.
- `count_0` / `percentage_0` — zero values for id-columns.
- `distinct_values` — number of unique non-null values for id-columns.
- `min`, `q05`, `q25`, `median`, `q75`, `q95`, `max` — quantile summaries for birth-date fields.

# Visualise the results

`tablePerson()` formats the `summarised_result` into a publication-ready table.
The `type` argument accepts `"gt"` (default), `"reactable"`, or `"datatable"`.

```{r, warning=FALSE}
tablePerson(result = result, type = "gt")
```

The column headers are driven by the CDM name and the estimate combinations are
rendered as human-readable labels (e.g. `N (%)`, `Median [Q25 - Q75]`,
`Missing N (%)`).

# Disconnect

```{r}
cdmDisconnect(cdm = cdm)
```
