---
title: "Summarise concept id counts"
output: 
  html_document:
    pandoc_args: [
      "--number-offset=1,0"
      ]
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{summarise_concept_id_counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In this vignette, we will explore the *OmopSketch* functions designed to summarise concept use in clinical OMOP tables. Specifically, there are two key functions:

-   `summariseConceptIdCounts()`: counts standard concept IDs and their associated source concept IDs in one or more clinical tables.

-   `tableConceptIdCounts()`: displays concept count results in a formatted table.

## Create a mock cdm

Let's see an example of these functions. To start with, we will load essential packages and create a mock CDM using the R package [omock](https://ohdsi.github.io/omock/).

```{r, warning=FALSE}
library(OmopSketch)
library(dplyr)
library(omock)

cdm <- mockCdmFromDataset(datasetName = "GiBleed", source = "duckdb")

cdm
```

# Summarise concept level counts

We now use the `summariseConceptIdCounts()` function from the **OmopSketch** package to retrieve counts for each standard concept ID and name, together with the associated source concept ID, source concept name, standard vocabulary, and source vocabulary.

```{r, warning=FALSE}
summariseConceptIdCounts(cdm = cdm, omopTableName = "drug_exposure") |>
  select(group_level, variable_name, variable_level, estimate_name, estimate_value, additional_name, additional_level) |>
  glimpse()
```

By default, the function returns the number of records (`estimate_name == "count_records"`) for each standard concept ID. To count distinct subjects instead, set `countBy = "person"`. To return both record and subject counts, use `countBy = c("record", "person")`.

```{r, warning=FALSE}
summariseConceptIdCounts(
  cdm = cdm,
  omopTableName = "drug_exposure",
  countBy = c("record", "person")
) |>
  select(variable_name, estimate_name, estimate_value)
```

Further stratification can be applied using the `interval`, `sex`, and `ageGroup` arguments. The `interval` argument supports `"overall"` (no time stratification), `"years"`, `"quarters"`, or `"months"`. Age groups and time intervals are assigned using the clinical record start date.

```{r, warning=FALSE}
summariseConceptIdCounts(
  cdm = cdm,
  omopTableName = "condition_occurrence",
  countBy = "person",
  interval = "years",
  sex = TRUE,
  ageGroup = list("<=50" = c(0, 50), ">50" = c(51, Inf))
) |>
  select(group_level, strata_level, variable_name, estimate_name, additional_level) |>
  glimpse()
```

When `inObservation = TRUE`, results are stratified by whether each record occurred within the subject's observation period.

```{r, warning=FALSE}
summariseConceptIdCounts(
  cdm = cdm,
  omopTableName = "condition_occurrence",
  countBy = "record",
  inObservation = TRUE
) |>
  select(variable_name, strata_name, strata_level, estimate_name, estimate_value) |>
  glimpse()
```

We can also filter the clinical table to a specific time window by setting the `dateRange` argument. Only records with a start date inside the date range are included.

```{r}
summarisedResult <- summariseConceptIdCounts(
  cdm = cdm,
  omopTableName = "condition_occurrence",
  dateRange = as.Date(c("1990-01-01", "2010-01-01"))
)
summarisedResult |>
  settings() |>
  glimpse()
```

You can restrict concept counts to a subset of subjects with the `sample` argument. Provide an integer to randomly select that many `person_id`s from the `person` table, or provide the name of a cohort table in the CDM to restrict counts to its `subject_id`s.

```{r}
summariseConceptIdCounts(
  cdm = cdm,
  omopTableName = "condition_occurrence",
  sample = 50
) |>
  select(group_level, variable_name, estimate_name) |>
  glimpse()
```

## Display the results

Concept counts can be displayed using `tableConceptIdCounts()`. By default, it generates an interactive [reactable](https://glin.github.io/reactable/) table, but [DT](https://rstudio.github.io/DT/) datatables are also supported.

```{r, warning=FALSE}
result <- summariseConceptIdCounts(
  cdm = cdm,
  omopTableName = "measurement",
  countBy = "record"
)
tableConceptIdCounts(result = result, type = "reactable")
```

```{r}
tableConceptIdCounts(result = result, type = "datatable")
```

The `display` argument in `tableConceptIdCounts()` controls which concept columns are shown. The default is `display = "overall"`, which shows both standard and source concept information.

```{r}
tableConceptIdCounts(result = result, display = "overall")
```

If `display = "standard"`, the table shows only standard concept ID, standard concept name, and standard vocabulary.

```{r}
tableConceptIdCounts(result = result, display = "standard")
```

If `display = "source"`, the table shows only source concept ID, source concept name, and source vocabulary.

```{r}
tableConceptIdCounts(result = result, display = "source")
```

If `display = "missing source"`, the table shows standard concepts for records that are missing a corresponding source concept ID.

```{r}
tableConceptIdCounts(result = result, display = "missing source")
```

If `display = "missing standard"`, the table shows source concepts for records that are missing a mapped standard concept ID.

```{r}
tableConceptIdCounts(result = result, display = "missing standard")
```

## Display the most frequent concepts

You can use the `tableTopConceptCounts()` function to display the most frequent concepts in an OMOP CDM table. By default, the function returns a [gt](https://gt.rstudio.com/) table, but other formats supported by `visOmopResults::tableType()` can also be used.

```{r, warning=FALSE}
result <- summariseConceptIdCounts(
  cdm = cdm,
  omopTableName = "drug_exposure",
  countBy = "record"
)
tableTopConceptCounts(result = result, type = "gt")
```

### Customising the number of top concepts

By default, the function shows the top 10 concepts within each table and stratum. You can change this using the `top` argument:

```{r}
tableTopConceptCounts(result = result, top = 5)
```

### Choosing the count type

If your summary includes both record and subject counts, you must specify which type to rank by using the `countBy` argument:

```{r, warning=FALSE}
result <- summariseConceptIdCounts(
  cdm = cdm,
  omopTableName = "drug_exposure",
  countBy = c("record", "person")
)
tableTopConceptCounts(result = result, countBy = "person")
```

# Disconnect from CDM

Finally, disconnect from the mock CDM.

```{r}
cdmDisconnect(cdm = cdm)
```
